A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-22 03-semantics.tex
Conserta if-else na gramática da Seção 3.2
Adiciona teorema da terminação no fim do apêndice
Correção Rodrigo e remoção do "Problema"
Revisão Seção 3-5

In this section, we introduce a reduced syntax for \CEU and propose an
operational semantics to formally describe the behavior of its programs.
We describe a small synchronous kernel highlighting the peculiarities of \CEU,
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language,
leaving out side-effects and system calls (which behave like in conventional
imperative languages).

@@

The grammar below defines the syntax of a subset of \CEU that is
sufficient to describe all semantic peculiarities of the language.
\bgroup
\def\lbl#1{\qquad\text{\emph{#1}}}%
\newdimen\X
\X=-1\jot
\begin{alignat*}{2}
  p\Coloneqq
      &\enspace\ceu{\Mem(\Id)}
      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitExt(\Id)}
      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitInt(\Id)}
      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\EmitInt(\Id)}
      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\Break}
      &&\lbl{loop escape}\\[\X]
      %%
  \mid&\enspace\ceu{\Ifelse{b}{p_1}{p_2}}
      &&\lbl{conditional}\\[\X]
      %%
  \mid&\enspace\ceu{p1\,;\,p2}
      &&\lbl{sequence}\\[\X]
      %%
  \mid&\enspace\ceu{\Loop p_1}
      &&\lbl{repetition}\\[\X]
      %%
  \mid&\enspace\ceu{\Every{\Id}\ p_1}
      &&\lbl{event iteration}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\And p_2}
      &&\lbl{par/and}\\[\X]
      %%
  \mid&\enspace\ceu{p_2\Or p_2}
      &&\lbl{par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\Fin p}
      &&\lbl{finalization}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtLoop p_2}
      &&\lbl{unwinded loop}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtAnd\ p_2}
      &&\lbl{unwinded par/and}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtOr\ p_2}
      &&\lbl{unwinded par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\CanRun(n)}
      &&\lbl{can run on stack level~$n$}\\[\X]
      %%
  \mid&\enspace\ceu{\Nop}
      &&\lbl{terminated program}
\end{alignat*}
\egroup

@@

Most expressions in the abstract language are mapped to
their counterparts in the concrete language.  The exceptions are the
finalization block~$\ceu{\Fin{p}}$ and the \texttt{@}-expressions, which are
internal expressions that result from the
expansion of awaits, emits, and loops by the transition rules to be discussed.

Regarding other mismatches between the concrete and abstract languages, the concrete
\code{await} and \code{emit} primitives support communication of values between
them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an $\ceu{\EmitInt}$ and access it after the
corresponding $\ceu{\AwaitInt}$.
%
Also, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete language is
equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract language.
In the concrete language, $A$ and~$C$ execute in sequence, and
the finalization code~$B$ is implicitly suspended waiting for~$C$
termination.
In the abstract language, ``$\ceu{\Fin B}$'' suspends forever when reached (it is
an awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute~$C$ in parallel, whose
termination aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to
execute (by the semantic rules to be discussed).

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way
that at most one transition is possible at any time, resulting in deterministic
reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~
%
\rc{Na BNF o $p$ não deveria ser $\P$~(maiúsculo) tbm?}
%
$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
nonnegative integers representing the current stack level,
and~
%
\rc{Checar: Acho que o conjunto $\E$~não foi previamente definido em
  lugar nenhum}
%
$e,e'\in\E\cup\{\nil\}$
are the events emitted before and after the
transition (both being possibly the empty event~$\nil$).

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the emitted event
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a (possibly) modified program~$p'$, at stack
level~0, and with no event emitted~($\nil$):
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

@@

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level:
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[2\jot]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$\ceu{p=\Nop}\vee\isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

Rule \R{pop} only decreases the stack level, not affecting the
program, and only applies if the program is blocked (see function~$\isblocked$ in
Figure~\ref{fig.isblocked}).
This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal
reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.

At the beginning of the reaction, an external event is emitted, which
will trigger rule \R{push}, which will immediately raise the stack level
to~1.
At the end of the reaction, the program will block or terminate and
successive applications of
rule~\R{pop} will eventually lead to a description containing this
same program at stack level~0.
(Rule \R{pop} is the only rule that decreases the stack level.)

@@

A complete reaction consists of the following series of transitions:
\begin{align*}
  \<p,0,e_\ext>\outpush\<p_1,1,\nil>
  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
  \null\nst[*]\null\outpop\<p',0,\nil>\,.
\end{align*}
%
%-
% \begin{align*}
% a) &\5\5
%     \LL p,0,ext \RR
%         \1\xrightarrow[out]{push}\1
%     \LL q,1,\epsilon \RR
% \\
% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
%     \LL r,i,e \RR
%         \1\xrightarrow[out]\1
%     \LL s,j,\epsilon \RR \1]*
% \\
% c) &\5\5 \1\xrightarrow[in]{*}\1
%     \LL t,k,\epsilon \RR
%         \1\xrightarrow[out]{pop}\1
%     \LL u,0,\epsilon \RR
% \end{align*}
%-
%
First, a~$\outpush$ starts a nested reaction at level~1.
Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
single~$\out$ transition (stack operation) takes place.
Finally, a last~$\outpop$ transition decrements the
stack level to~0 and terminates the reaction.

@@

A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory access (rule
\R{mem}).
An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a $\ceu{\CanRun(n)}$ which
can only resume at level~$n$ (rule \R{emit-int}).
Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$ causes
rule \R{push} to execute at the outer level, creating a new level~$n+1$ on
the stack.
Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
transit (rule~\R{can-run}), providing the desired stack-based semantics for internal events.

Proceeding to compound expressions, the rules for conditionals and sequences
are straightforward:
%
\begin{gather*}
  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{if-true}\\[2\jot]
  %%
  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\Ifelse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
  \DisplayProof
  \Rtag{if-false}\\[2\jot]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
  \DisplayProof
  \Rtag{seq-adv}
\end{gather*}
\vskip-\belowdisplayskip
\vskip-\abovedisplayskip
\vskip2\jot
\begin{align*}
  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[2\jot]
  %%
  \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
\end{align*}

@@

Given that our semantics focuses on control, rules \R{if-true} and
\R{if-false} are the only to query~$\ceu{\Mem}$ expressions.
%
Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression.
%
%Although the value here is arbitrary, it is unique in a reaction, because a
%given expression can execute only once within it (remember that $loops$ must
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

Finally, a~$\ceu{\Break}$ in one of the sides in parallel escapes the closest
enclosing~$\ceu{\Loop}$, properly aborting the other side by applying the~$\clear$
function:
\begin{gather*}
  \hskip-.5em
  \<\ceu{{\Break}\AtAnd{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
  \Rtag{and-brk1}\\[2\jot]
  %%
  \hskip-.5em
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtAnd{\Break}},n,\nil>
    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
  \DisplayProof
  \Rtag{and-brk2}\\[2\jot]
  %%
  \<\ceu{{\Break}\AtOr{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
  \Rtag{or-brk1}\\[2\jot]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtOr{\Break}},n,\nil>
    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
  \DisplayProof
  \Rtag{or-brk2}
\end{gather*}

@@

A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$, $\ceu{\Fin}$,
and~$\ceu{\CanRun}$ expressions in parallel trails.
%
If no trails are blocked in~$\ceu{\CanRun}$ expressions, it means that the program
cannot advance in the current reaction.
%
However, $\ceu{\CanRun}$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \R{pop}).
%
\rc{Não entendi essa discussão}

>>>>

In this section, we introduce a reduced syntax of \CEU and propose an
operational semantics to formally describe the behavior of programs.
We describe a small synchronous kernel highlighting the peculiarities of \CEU,
in particular, the stack-based execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language,
leaving out side-effects and system calls (which behave like in conventional
imperative languages).

@@

The grammar below defines the syntax of a subset of \CEU that is
sufficient to describe all semantic peculiarities of the language.
\bgroup
\def\lbl#1{\quad\text{\emph{#1}}}%
\newdimen\X
\X=-1\jot
\begin{alignat*}{2}
  p\Coloneqq
      &\enspace\ceu{\Mem(\Id)}
      &&\lbl{any memory access to~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitExt(\Id)}
      &&\lbl{await external event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\AwaitInt(\Id)}
      &&\lbl{await internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\EmitInt(\Id)}
      &&\lbl{emit internal event~``$\ceu\Id$''}\\[\X]
      %%
  \mid&\enspace\ceu{\Break}
      &&\lbl{loop escape}\\[\X]
      %%
  \mid&\enspace\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}
      &&\lbl{conditional}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\,;\,p_2}
      &&\lbl{sequence}\\[\X]
      %%
  \mid&\enspace\ceu{\Loop p_1}
      &&\lbl{repetition}\\[\X]
      %%
  \mid&\enspace\ceu{\Every{\Id}\ p_1}
      &&\lbl{event iteration}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\And p_2}
      &&\lbl{par/and}\\[\X]
      %%
  \mid&\enspace\ceu{p_2\Or p_2}
      &&\lbl{par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\Fin p}
      &&\lbl{finalization}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtLoop p_2}
      &&\lbl{unwinded loop}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtAnd\ p_2}
      &&\lbl{unwinded par/and}\\[\X]
      %%
  \mid&\enspace\ceu{p_1\AtOr\ p_2}
      &&\lbl{unwinded par/or}\\[\X]
      %%
  \mid&\enspace\ceu{\CanRun(n)}
      &&\lbl{can run on stack level~$n$}\\[\X]
      %%
  \mid&\enspace\ceu{\Nop}
      &&\lbl{terminated program}
\end{alignat*}
\egroup

@@

Most expressions in the abstract language are mapped to their counterparts in
the concrete language.
The exceptions are the finalization block~$\ceu{\Fin{p}}$ and the
\texttt{@}-expressions which result from expansions in the transition rules to
be presented.

Regarding mismatches between the concrete and abstract languages, the concrete
\code{await} and \code{emit} primitives support communication of values between
them, e.g., an ``\code{emit a(10)}'' awakes a ``\code{v=await a}'' setting
variable~\code{v} to~10.
To reproduce this functionality in the formal semantics, we can use a shared
variable to hold the value of an $\ceu{\EmitInt}$ and access it after the
corresponding $\ceu{\AwaitInt}$ awakes.
%
Also, a ``\code{finalize $A$ with $B$ end; $C$}'' in the concrete language is
equivalent to ``\ceu{A;\;((\Fin{B})\ \Or\ C)}'' in the abstract language.
In the concrete language, $A$ and~$C$ execute in sequence, and
the finalization code~$B$ is implicitly suspended waiting for~$C$
termination.
In the abstract language, ``$\ceu{\Fin B}$'' suspends forever when reached (it is
an awaiting expression that never awakes).
Hence, we need an explicit \code{or} to execute~$C$ in parallel, whose
termination aborts ``$\ceu{\Fin B}$'', which finally causes~$B$ to
execute (by the semantic rules to be discussed).

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are designed in such a way
that at most one transition is possible at any time, resulting in deterministic
reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are non-negative
integers representing the current stack level, and~$e,e'\in\E\cup\{\nil\}$ are
the events emitted before and after the transition (both being possibly the
empty event~$\nil$).

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a (possibly) modified program~$p'$, at stack
level~0, and with no event emitted~($\nil$):
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

@@

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions which only
apply to the program as a whole, and are the only to manipulate the stack
level:
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[2\jot]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$\ceu{p=\Nop}\vee\isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

Rule \R{pop} only decreases the stack level, not affecting the
program, and only applies if the program is blocked (see function~$\isblocked$ in
Figure~\ref{fig.isblocked}).
This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal
reaction completes and blocks in the current stack level.

At the beginning of the reaction, an external event is emitted, which
will trigger rule \R{push}, which will immediately raise the stack level
to~1.
At the end of the reaction, the program will block or terminate and
successive applications of
rule~\R{pop} will eventually lead to a description containing this
same program at stack level~0.

@@

A complete reaction consists of a series of transitions:
\begin{align*}
  \<p,0,e_\ext>\outpush\<p_1,1,\nil>
  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
  \null\nst[*]\null\outpop\<p',0,\nil>\,.
\end{align*}
%
%-
% \begin{align*}
% a) &\5\5
%     \LL p,0,ext \RR
%         \1\xrightarrow[out]{push}\1
%     \LL q,1,\epsilon \RR
% \\
% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
%     \LL r,i,e \RR
%         \1\xrightarrow[out]\1
%     \LL s,j,\epsilon \RR \1]*
% \\
% c) &\5\5 \1\xrightarrow[in]{*}\1
%     \LL t,k,\epsilon \RR
%         \1\xrightarrow[out]{pop}\1
%     \LL u,0,\epsilon \RR
% \end{align*}
%-
%
First, a~$\outpush$ starts a nested reaction at level~1.
Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
single~$\out$ transition (stack operation) takes place.
Finally, a last~$\outpop$ transition decrements the
stack level to~0 and terminates the reaction.

@@

A $\ceu{\Mem}$ operation becomes a $\ceu{\Nop}$ which indicates the memory
access (rule \R{mem}).
An $\ceu{\EmitInt(id)}$ generates an event $\ceu{\Id}$ and transits to a
$\ceu{\CanRun(n)}$ which can only resume at level~$n$ (rule \R{emit-int}).
Since all~$\nst$ rules can only transit with $e=\nil$, an $\ceu{\EmitInt}$
inevitably causes rule \R{push} to execute at the outer level, creating a new
level~$n+1$ on the stack.
Also, with the new stack level, the resulting $\ceu{\CanRun}(n)$ itself cannot
transit yet (rule~\R{can-run}), providing the desired stack-based semantics for
internal events.

Proceeding to compound expressions, the rules for conditionals and sequences
are straightforward:
%
\begin{gather*}
  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{if-true}\\[2\jot]
  %%
  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
  \DisplayProof
  \Rtag{if-false}\\[2\jot]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
  \DisplayProof
  \Rtag{seq-adv}
\end{gather*}
\vskip-\belowdisplayskip
\vskip-\abovedisplayskip
\vskip2\jot
\begin{align*}
  \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[2\jot]
  %%
  \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
\end{align*}

@@

Given that our semantics focuses on control, rules \R{if-true} and
\R{if-false} are the only to query~$\ceu{\Mem}$ expressions.
%
Function~$\eval$ evaluates a given~$\ceu{\Mem}$ expression to a boolean value.
%
%Although the value here is arbitrary, it is unique in a reaction, because a
%given expression can execute only once within it (remember that $loops$ must
%contain $awaits$ which, from rule \textbf{await}, cannot awake in the same
%reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

Finally, a~$\ceu{\Break}$ in one of the sides in parallel escapes the closest
enclosing~$\ceu{\Loop}$, properly aborting the other side with the~$\clear$
function:
\begin{gather*}
  \hskip-.5em
  \<\ceu{{\Break}\AtAnd{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
  \Rtag{and-brk1}\\[2\jot]
  %%
  \hskip-.5em
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtAnd{\Break}},n,\nil>
    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
  \DisplayProof
  \Rtag{and-brk2}\\[2\jot]
  %%
  \<\ceu{{\Break}\AtOr{q}},n,\nil>\nst\<\ceu{\clear(q);\Break},n,\nil>
  \Rtag{or-brk1}\\[2\jot]
  %%
  \AxiomC{$\isblocked(p,n)$}
  \UnaryInfC{$\<\ceu{p\AtOr{\Break}},n,\nil>
    \nst\<\ceu{\clear(p);\Break},n,\nil>$}
  \DisplayProof
  \Rtag{or-brk2}
\end{gather*}

@@

A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$,
$\ceu{\Every}$, $\ceu{\Fin}$, and~$\ceu{\CanRun}$ expressions in parallel
trails.
%
Then, if none of the trails is blocked in~$\ceu{\CanRun}$ expressions, it means
that the program cannot advance in the current reaction.
%
However, $\ceu{\CanRun}$ expressions can still resume at lower stack indexes
and will eventually resume in the current reaction (see rule \R{pop}).

\begin{figure}[h]
\small
\begin{gather*}
  \boxed{
    \begin{align*}
      %%
      %%-
      \intertext{\llap{(i)~}Function~$\bcast$:}
      %%-
      %%
      \bcast(\ceu{\AwaitExt(e)},e)
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \bcast(\ceu{\AwaitInt(e)},e)
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \bcast(\ceu{\Every{e}\ {p}},e)
      &=\ceu{p;\,\Every{e}\ {p}}\\[-1\jot]
      %%
      \bcast(\ceu{\CanRun(n)},e)
      &=\ceu{\CanRun(n)}\\[-1\jot]
      %%
      \bcast(\ceu{\Fin{p}},e)
      &=\ceu{\Fin{p}}\\[-1\jot]
      %%
      \bcast(\ceu{p;\,q})
      &=\ceu{\bcast(p,e);\,q}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtLoop{q}},e)
      &=\ceu{\bcast(p,e)\AtLoop{q}}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtAnd{q}},e)
      &=\ceu{{\bcast(p,e)}\AtAnd{\bcast(q,e)}}\\[-1\jot]
      %%
      \bcast(\ceu{p\AtOr{q}},e)
      &=\ceu{{\bcast(p,e)}\AtOr{\bcast(q,e)}}\\[-1\jot]
      %%
      bcast(\_,e)
      &=\_\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
      \\[1\jot]
      %%
      %%-
      \intertext{\llap{(ii)~}Predicate~$\isblocked$:}
      %%-
      %%
      \isblocked(\ceu{\AwaitExt(e)},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\AwaitInt(e)},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\Every{e}\ {p}},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{\CanRun(m)},n)
      &=(n>m)\\[-1\jot]
      %%
      \isblocked(\ceu{\Fin{p}},n)
      &=\mathit{true}\\[-1\jot]
      %%
      \isblocked(\ceu{p;\,q},n)
      &=\isblocked(p,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtLoop{q}},n)
      &=\isblocked(p,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtAnd{q}},n)
      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
      %%
      \isblocked(\ceu{p\AtOr{q}},n)
      &=\isblocked(p,n)\land\isblocked(q,n)\\[-1\jot]
      %%
      \isblocked(\_,n)
      &=\mathit{false}\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
        \\[1\jot]
      %%
      %%-
      \intertext{\llap{(iii)~}Function~$\clear$:}
      %%-
      %%
      \clear(\ceu{\AwaitExt(e)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\AwaitInt(e)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\Every{e}\ p})
      %%
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\CanRun(n)})
      &=\ceu{\Nop}\\[-1\jot]
      %%
      \clear(\ceu{\Fin{p}})
      &=p\\[-1\jot]
      %%
      \clear(\ceu{p;\,q})
      &=\clear(p)\\[-1\jot]
      %%
      \clear(\ceu{p\AtLoop{q}})
      &=\clear(p)\\[-1\jot]
      %%
      \clear(\ceu{p\AtAnd{q}})
      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
      %%
      \clear(\ceu{p\AtOr{q}})
      &=\ceu{\clear(p);\,\clear(q)}\\[-1\jot]
      %%
      \clear(\_)
      &=\bot\enspace
        (\ceu{\Mem},\ceu{\EmitInt},\ceu{\Break},\\[-1\jot]
      &\quad\ceu{\IfElse{}{}},\ceu{\Loop},\ceu{\And},\ceu{\Or},\ceu{\Nop})
    \end{align*}}
\end{gather*}
\vskip-2\belowdisplayskip
\caption{%
  (i)~Function~$\bcast$ awakes awaiting trails matching the event by
  converting~$\ceu{\protect\AwaitExt}$ and~$\ceu{\protect\AwaitInt}$
  to~$\ceu{\protect\Nop}$ expressions, and by unwinding $\ceu{\protect\Every}$
  expressions.
  %%
  \enspace(ii)~Predicate~$\isblocked$ is true only if all branches in parallel
  are blocked waiting for events, for finalization clauses, or for certain
  stack levels.
  %%
  \enspace(iii)~Function~$\clear$ extracts~$\ceu{\protect\Fin}$ expressions in
  parallel and put their bodies in sequence.
}
\label{fig.bcast}
\label{fig.isblocked}
\label{fig.clear}
\end{figure}

### 2018-02-22 04-related.tex
Revisão Seção 3-5

\CEU follows the lineage of imperative synchronous languages initiated by
Esterel~\cite{esterel.ieee91}.
These languages typically define time as a discrete sequence of logical
``ticks'' in which multiple simultaneous input events can be
active~\cite{ceu.tecs17}.
%
The presence of multiple inputs requires careful static analysis to detect and
reject programs with \emph{causality cycles} and \emph{schizophrenia problems}%
~\cite{esterel.constructive}.
%
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events, which is a prerequisite for the concurrency checks that
enable safe shared-memory concurrency.

Also, in most synchronous languages, the behavior of internal and external
events is equivalent.
In \CEU, internal events introduce stack-based micro reactions within external
reactions, providing more fine-grained control for intra-reaction execution.
%
This allows for memory-bounded subroutines that execute multiple times in the
same external reaction.
%
The synchronous languages Statecharts~\cite{statecharts.variants} and
Statemate~\cite{statecharts.statemate} also distinguish internal from external
events.
In the former, \emph{``reactions to external and internal events (...) can be
sensed only after completion of the step''}.
In the latter, \emph{``the receiving state (of the internal event) acts here as
a function''}.
Although the descriptions suggest a stack-based semantics, we are not aware of
formalizations for these ideas for a deeper comparison with \CEU.

Like \CEU, many synchronous languages%
~\cite{rp.rc,wsn.protothreads,wsn.sol,rp.synchc,rp.pretc}
also rely on lexical scheduling to preserve determinism.
%
In contrast, in Esterel, the execution order for operations within a reaction
is non-deterministic: \emph{``if there is no control dependency, as in
\code{(call~f1()~||~call~f2())},
the order is unspecified and it would be an error to rely on
it''}~\cite{esterel.primer}.
%
For this reason, Esterel, does not support shared-memory concurrency:
\emph{``if a variable is written by some thread, then it can neither be read
nor be written by concurrent threads''}~\cite{esterel.primer}.

Regarding the integration with the $C$ language-based environment, \CEU
supports a finalization mechanism for external resources.
%
In addition, \CEU also tracks pointers representing resources that cross $C$
boundaries and forces the programmer to provide associated finalizers.

>>>>

\CEU follows the lineage of imperative synchronous languages initiated by
Esterel~\cite{esterel.ieee91}.
These languages typically define time as a discrete sequence of logical
``ticks'' in which multiple simultaneous input events can be
active~\cite{ceu.tecs17}.
%
The presence of multiple inputs requires careful static analysis to detect and
reject programs with \emph{causality cycles} and \emph{schizophrenia problems}%
~\cite{esterel.constructive}.
%
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events, which is a prerequisite for the concurrency checks that
enable safe shared-memory concurrency, as discussed in
Section~\ref{sec.ceu.shared}.

In most synchronous languages, the behavior of external and internal events is
equivalent.
However, in \CEU, internal events introduce stack-based micro reactions within
external reactions, providing more fine-grained control for intra-reaction
execution.
%
This allows for memory-bounded subroutines that can execute multiple times
during the same external reaction.
%
The synchronous languages Statecharts~\cite{statecharts.variants} and
Statemate~\cite{statecharts.statemate} also distinguish internal from external
events.
In the former, \emph{``reactions to external and internal events (...) can be
sensed only after completion of the step''}.
In the latter, \emph{``the receiving state (of the internal event) acts here as
a function''}.
Although the descriptions suggest a stack-based semantics, we are not aware of
formalizations for these ideas for a deeper comparison with \CEU.

Like \CEU, many other synchronous languages%
~\cite{rp.rc,wsn.protothreads,wsn.sol,rp.synchc,rp.pretc}
also rely on lexical scheduling to preserve determinism.
%
In contrast, in Esterel, the execution order for operations within a reaction
is non-deterministic: \emph{``if there is no control dependency, as in
\code{(call~f1()~||~call~f2())},
the order is unspecified and it would be an error to rely on
it''}~\cite{esterel.primer}.
%
For this reason, Esterel, does not support shared-memory concurrency:
\emph{``if a variable is written by some thread, then it can neither be read
nor be written by concurrent threads''}~\cite{esterel.primer}.

Regarding the integration with $C$ language-based environments, \CEU
supports a finalization mechanism for external resources.
%
In addition, \CEU also tracks pointers representing resources that cross $C$
boundaries and forces the programmer to provide associated finalizers.
%
As far as we know, this extra safety level is unique to \CEU.
